Robust Quantum Optimal Control

### 2020-07-20 paper/paper.tex
second pass through section 3

Numerical quantum control techniques have developed in parallel
with the analytic frameworks. Numerical quantum control
techniques integrate the time-dependent Schroedinger
equation and use sensitivity analysis to optimize target metrics
of the quantum gate, such as fidelity
\cite{leung2017speedup,  goerz2019krotov, doria2011optimal,
  abdelhafez2019gradient, machnes2015gradient, leng2019robust}.
These techniques typically employ zeroth- or first-order
optimizers and make no garauntees of optimality.
Further, they formulate the quantum control
problem as unconstrained or use
projective gradient methods \cite{machnes2015gradient}.
The latter approach relies on constructing
bases for constraint spaces--which is not always feasible--and
may hinder convergence.

@@

%% SECTION 2
\section{QOC + AL-iLQR}
(QOC Problem Statement) Here we introduce the notation
we will use throughout the paper,
review the quantum optimal control problem statement,
and introduce the trajectory optimization framework.
Quantum optimal control concerns the evolution of
a state $\ket{\psi(t)}$ governed by the time-dependent
Schroedinger equation
\begin{align}
  \partial_{t} \ket{\psi(t)} &= -\frac{i}{\hbar} H(\textbf{u}(t), t) \ket{\psi(t)}
\end{align}
The evolution is sometimes cast with $\rho(t) = \ket{\psi(t)}\bra{\psi(t)}$
and the Lindblad master equation to model the decoherence of the
state explicitly. The Hamiltonian
depends arbitrarily on the controls $\textbf{u}(t)$.

Numerical quantum optimal control techniques make
the problem tractable by discretizing the problem into $N$
time steps. Typical integration techniques rely on approximating
the analytical solution to the time-dependent Schroedinger equation
or using explicit integration methods, e.g. Runge-Kutta, of the form
\begin{align}
  \ket{\psi_{k + 1}} \approx \ket{\psi_{k}} + \partial_{t} \ket{\psi_{k}} \cdot \Delta t_{k}
\end{align}

Quantum optimal control seeks the control
parameters that minimize a functional $J(u, \ket{\psi_{N}(u)})$.
In the simplest case the functional
$J = 1 - {\lvert \braket{\psi_{f} \lvert \psi_{N}(u)} \rvert}^{2}$
which is the infidelity between the inital state evolved
to the final time step ($\ket{\psi_{N}(u)}$)
and the target state ($\ket{\psi_{f}}$). In general
$J$ is a linear combinaion of cost functions on $\ket{\psi_{N}(u)}$, e.g.
forbidden-state occupation, as well as
cost functions on $u$, e.g. the norm of the control amplitudes
\cite{leung2017speedup}.

(AL-iLQR Problem Statement) The trajectory optimization
literature solves a more general class of non-linear programs that resemble
the quantum optimal control problem. The quantum optimal control
problem is a specific case of the linear quadratic regulator (LQR).
LQR is so called because the dynamics are linear in the state and
the cost function is quadratic in the state. In the LQR formulation
the same functional is evaluated at each time step
\begin{equation}
  \begin{aligned}
    J_{\textrm{LQR}} = \tilde{x}_{N}^{T} Q_{N} \tilde{x}_{N}
    &+ \sum_{k = 0}^{N - 1} \tilde{x}_{k}^{T} Q_{k} \tilde{x}_{k} + u_{k}^{T} R_{k} u_{k}
  \end{aligned}
\end{equation}
where $\tilde{x}_{k} = x_{k} - x_{f}$ is the difference between the current
and final state and $Q_{k}, R_{k}$ are matrices that define the penalty metric.

The advantage of the LQR formulation
is that there exists a dynamic programming algorithm to compute the
optimal update to the controls $u_{k}$ at each time step.
This algorithm employs the Ricatti recursion (see Appendix).

In order to incorporate constraints into the problem we employ
the augmented lagrangian method. This method adaptively
updates the penalty of constraints
%% LEFT OFF HERE

Trajectory optimization gives us gaurantees
about our updates via Ricatti recursion
and allows us to put constraints on our cost functions.
%% This is e.q. 2 from ALTRO paper
\begin{equation}
\begin{aligned}[b]
  \min_{U:N-1 \times m} &\mathcal{L}_{N}(x_{N}, \lambda_{N}, \mu_{N})\\
  &+ \sum_{k = 1}^{N - 1} \mathcal{L}_{k}(x_{k}, u_{k}, \lambda_{k}, \mu_{k})\\
\end{aligned}
\end{equation}
\begin{equation}
\begin{aligned}[b]
  \mathcal{L}_{k}(x_{k}, &u_{k}, \lambda_{k}, \mu_{k}) =\\
  &(\lambda_{k} + \frac{1}{2}I_{\mu_{k}} c_{k}(x_{k}, u_{k}))^{T} c_{k}(x_{k}, u_{k})\\
  &+ (x_{k} - x_{f})^{T} Q (x_{k} - x_{f}) + u_{k}^{T} R u_{k}\\
\end{aligned}
\end{equation}
The important point is that there is an update step (e.q. 17 from ALTRO paper) where we send $\lambda \rightarrow \infty$
and get all of the nice convergence properties. The weights are adjusted dynamically between iterations until
all of our constraints are satisfied. The Markovian decision structure of the problem allows
us to apply differentiable dynamic programming to garauntee that the update for each control $U_{k}$ is
opimal, as apposed to the greedy updates of first-order optimizers like the naive gradient descent.

%% TODO: (?) Figure comparing update procedures

%% SECTION 3
\section{QOC on the Fluxonium}
(Fluxonium Device) In the two-level
approximation we have
\begin{align}
  H/h &= \omega_{q} \frac{\sigma_{z}}{2} + A(\Phi_{ext}) \frac{\sigma_{x}}{2}
\end{align}
This approximation is good up to the avoided crossing at $\Phi_{\textrm{ext}} = 0.35 \Phi_{0}$. We get $A$
by converting via $\bra{g} \hat{\phi} \ket{e}$.
We use the gates $X/2, Y/2, Z/2$. $X/2, Y/2$ are universal
up to an arbitrary Z rotation.
%% TODO: list device parameters

(Constraints) We want constraints on our pulses. We want pulses start and end at zero
for concatenation. We want pulses to have zero net flux to mitigate
hysteresis in flux bias lines.
%% TODO: See ref. 28 in Helin paper
We want the amplitude to be constrained
$\delta \Phi_{\textrm{ext}} \sim 0.06 \Phi_{0}$
so the two-level approximation stays valid.
We want our state to obey normalization conditions,
mitigating numerical error in simulaiton.
%% TODO: mention pulse smoothing via optimizing over 2nd derivative
%% of pulses
%% TODO: probably want to list equations for the constraints

($T_{1}$ and $T_{\phi}$ noise)
Dissipation to the thermal bath via longitudinal
relaxation ($T_{1}$) is an irreversible process
that results in information loss.
Pure dephasing ($T_{\phi}$) is a reversible process.
Dielectric loss, etc. contribute
to $T_{1}$. $1/f$ flux noise, etc. contribute to $T_{\phi}$
There is a tradeoff between $T_{1}$ and $T_{\phi}$. In the case of white
noise we have that the sum of the noise weights $W_{1}$ and $W_{\phi}$
is constant \cite{huang2020engineering}.
In our case $T_{\phi}$ is saturated at the flux
frustration point $\Phi_{\textrm{ext}} = 0.5 \Phi_{0}$
$T_{\phi}$ decreases as we move away from the flux
frustration point. Conversely, $T_{1}$ is at a minimum
around the flux frustration point $T_{1}(0.5 \Phi_{0}) = 0.315$ms,
and increases away from the flux frustration point
$T_{1}(0.43 \Phi_{0}) = 4.3$ms. Given the nature
of the decay processes and the tradeoff, we choose
to optimize the flux bias to have high time-averaged
$T_{1}$ times and employ some techniques to mitigate
$T_{\phi}$ dephasing.
%% TODO: See fig. 3 in Helin paper

%% SECTION 4
\section{Robustness to $T_{1}$-type Noise}
(Strategy) We seek to minimize the probability
that the qubit relaxes longitudinally.
To this end we augment the state vector $x_{k}$ with the
longitudinal relaxation probability
\begin{align}
  P_{1}(t_{k}) = \int_{0}^{t_{k}}
  \gamma_{1}(\Phi_{\textrm{ext}}(t)) dt
\end{align}
where $\gamma_{1} = T_{1}^{-1}$. Setting the target
longitudinal relaxation probability to 0 results in
a quadratic cost at each knot point
of the form ${\lvert P_{1}(t_{k}) \rvert}^{2}$.

@@

%% APPENDIX 2
\section{Measurements}
We measure $T_{1}$ using the standard experiment
and $T_{2}$ using te Ramsey experiment. We fit with splines
and the data looks like fig. 3 in Helin's paper \cite{zhang2020universal}.
We measure $\omega_{q}$ and $\sigma_{\omega_{q}}$ using X method.

>>>>

Numerical quantum control techniques have developed in parallel
with the analytic frameworks. Numerical quantum control
techniques integrate the time-dependent Schroedinger
equation and use sensitivity analysis to optimize target metrics
of the quantum gate, such as fidelity
\cite{leung2017speedup,  goerz2019krotov, doria2011optimal,
  abdelhafez2019gradient, machnes2015gradient, leng2019robust}.
These techniques typically employ zeroth- or first-order
optimizers and make no garauntees of optimality.
Further, they formulate the quantum control
problem as unconstrained or use
projective gradient methods to enforce constraints
\cite{machnes2015gradient}.
The latter approach relies on constructing
bases for constraint spaces--which is not always feasible--and
may hinder convergence.

@@

%% SECTION 2
\section{QOC + AL-iLQR}
(QOC Problem Statement) Here we introduce the notation
we will use throughout the paper,
review the quantum optimal control problem statement,
and introduce the trajectory optimization framework.
Quantum optimal control concerns the evolution of
a quantum state $\ket{\psi(t)}$ governed by the time-dependent
Schroedinger equation (TDSE)
\begin{align}
  \frac{d}{dt} \ket{\psi(t)} &= -\frac{i}{\hbar} H(\textbf{u}(t), t) \ket{\psi(t)}
\end{align}
The evolution is sometimes cast with the evolution
of a density matrix under the Lindblad master equation to
model the decoherence of the state explicitly. The Hamiltonian
depends arbitrarily on the controls ($\textbf{u}(t)$). The controls
are so called because they are the means the experimentalist has to
act on the system.

Numerical quantum optimal control techniques make
the problem tractable by discretizing the problem into $N$
time steps. Typical integration techniques for the TDSE include
approximating unitary propagators as well as explicit integration methods,
such as Runge-Kutta, of the form
\label{step-approx}
\begin{align}
  \ket{\psi_{k + 1}} \approx \ket{\psi_{k}} + \frac{d}{dt} \ket{\psi_{k}} \cdot \Delta t_{k}
\end{align}

Quantum optimal control seeks the control
parameters that minimize a functional $J(\textbf{u}(t))$.
In the simplest case the functional is
$J = 1 - {\lvert \braket{\psi_{f} \lvert \psi_{N}(\textbf{u}(t))} \rvert}^{2}$
the infidelity between the inital state evolved
to the final time step ($\ket{\psi_{N}(\textbf{u}(t))}$)
and the target state ($\ket{\psi_{f}}$). In general
$J$ is a linear combinaion of cost functions on the state, e.g.
forbidden-state occupation, as well as
cost functions on the controls, e.g. the norm of the control amplitudes
\cite{leung2017speedup}. Typical quantum optimal control
algorithms employ automatic differentiation
to compute first order information for the functional ($\nabla_{u} J(u)$).
They employ a first-order optimizer to minimize $J$ with respect to $u$.

(AL-iLQR Problem Statement) The trajectory optimization
literature solves a more general class of non-linear programs that resemble
the quantum optimal control problem. The quantum optimal control
problem is a specific case of the linear quadratic regulator (LQR).
LQR is so called because the dynamics are linear in the state and
the functional is quadratic in the state. In the LQR formulation
the same functional is evaluated at each time step
\begin{equation}
  J_{\textrm{iLQR}} = \tilde{x}_{N}^{T} Q_{N} \tilde{x}_{N}
  + \sum_{k = 0}^{N - 1} \tilde{x}_{k}^{T} Q_{k} \tilde{x}_{k} + u_{k}^{T} R_{k} u_{k}
\end{equation}
where $\tilde{x}_{k} = x_{k} - x_{f}$ is the difference between the state
at time step $k$ and final state, $u_{k}$ are the controls,
and $Q_{k}, R_{k}$ are matrices that define the penalty metric.
The state is propagated using a dynamics function similar to \ref{step-approx}
$x_{k + 1} = f(x_{k}, u_{k}, t_{k}, \Delta t_{k})$.
In the case of quantum optimal control $\ket{\psi_{k}} \subseteq x_{k}$
and $f$ encodes the TDSE dynamics.

The advantage of the LQR formulation
is that there exists a dynamic programming algorithm to compute the
optimal update to the controls ($u_{k}$) which minimizes the functional
($J_{\textrm{iLQR}, k}$) for each time step. This algorithm proceeds by deriving a
recurrence relation between time steps $k$ and $k + 1$ for the optimal
feedback law--known as the Ricatti recursion (see Appendix). The
iterative LQR (iLQR) algorithm computes $J_{\textrm{iLQR}}$
and applies the Ricatti recursion to all time steps on multiple
executions.

In order to incorporate constraints we employ
the augmented Lagrangian method. Constraints are contributions
to the functional of arbitrary form $c_{k}(x_{k}, u_{k})$ which are
zero-valued when the constraint is satisfied. The AL-iLQR
method updates the penalty multiplier between
iLQR executions. The penalty multiplier associated with the
constraint functional estimates the Lagrange multiplier of
the constraint. The functional now takes the form
\begin{equation}
  \begin{aligned}
    J_{\textrm{AL-iLQR}} = \ &(\lambda_{k} + \frac{1}{2}I_{\mu_{k}} c_{k}(x_{k}, u_{k}))^{T} c_{k}(x_{k}, u_{k})\\
    &+ J_{\textrm{iLQR}}
  \end{aligned}
\end{equation}
$\lambda_{k}$ is a Lagrange multiplier. $I_{\mu_{k}}$ is a penalty matrix
with $\mu_{k}$ along the diagonal. $i$ indicates the $i$-th constraint functional.
$\lambda_{k}$ and $\mu_{k}$ are updated after each augmented Lagrangian iteration according to
\begin{align}
  \lambda_{k_{i}}^{'} &= \lambda_{k_{i}} + \mu_{k_{i}} c_{k_{i}}(x_{k}^{*}, u_{k}^{*})\\
  \mu_{k_{i}}^{'} &= \phi \mu_{k_{i}}
\end{align}
where $x^{*}, u^{*}$ are the optimal state and controls from the iLQR execution
and $\phi$ is a hyperparameter. With this updated form of the cost
functional there still exists a recurrence relation to calculate the optimal control
updates, see \cite{howell2019altro}.

%% SECTION 3
\section{QOC on the Fluxonium}
(Fluxonium Device) In the following we study
the quantum optimal control problem on the fluxonium qubit.
In the two-level
approximation the system Hamiltonian takes the form
\label{hamiltonian}
\begin{align}
  H/h &= \omega_{q} \frac{\sigma_{z}}{2} + u(t) \frac{\sigma_{x}}{2}
\end{align}
$h$ is Planck's constant. The control parameter $u$ is experimentally
realized by modulating the flux 
threading the superconducting junction $\Phi_{\textrm{ext}}$.
The conversion is defined by
$u = 4 \pi \braket{g \lvert \hat{\phi} \rvert e} E_{L} \delta \Phi_{\textrm{ext}} / \Phi_{0}$
where $\delta \Phi_{\textrm{ext}} = \Phi_{\textrm{ext}} - 0.5 \Phi_{0}$ is the flux
offset from the flux frustration point.

In the following we compare the gates found with our methods
to the $X/2, Y/2, Z/2$ gates reported in \cite{zhang2020universal}
for the same device. Note that $Y/2$ and arbitrary $Z$ rotations
are sufficient for universal computation.
%% TODO: list device parameters

(Constraints) We now outline the constraints that we require each gate
to obey.
We require $u(t = 0) = u(t = t_{N}) = 0$.
This constraint ensures gates may be concatenated arbitrarily without
inducing AWG ringing due to high-frequency transitions.
Furthermore, we require $\int_{0}^{t_{N}} u(t) dt = 0$. This
constraint ensures the pulse has zero net flux, mitigating
the hysteresis ubiquitous in flux bias lines.
%% TODO: See ref. 28 in Helin paper
We require $-300 \textrm{MHz} \le u(t) \le 300 \textrm{MHz}$
to ensure the two-level approximation \ref{hamiltonian}
remains valid. Additionally, we require that each gate achieves
the desired state transition $\braket{\psi_{f} \lvert \psi_{N}} = 1$.
In addition to these constraints we penalize the norm
of the first and second derivatives of the control
to mitigate AWG ringing.

The equality and inequality constraints on $u$ are handled
with a bound constraint. $\int_{0}^{t_{k}} u(t) dt$ is added
to the augmented state vector $x_{k}$ by incorporating the
integration of $u_{k}$ into the dynamics function $f$.
Both the zero net flux and target quantum state constraint
are then handled by ensuring the augmented final state is
reached $x_{N} - x_{f} = 0$. The optimization is performed
over the second derivative of the controls $\frac{d^{2}}{dt^{2}} u(t)$
so that the second derivative, first derivative, zeroth derivative, and
integral of the controls may be penalized.
%% TODO: probably want to list equations for the constraints

($T_{1}$ and $T_{\phi}$ noise)
In addition to optimizing the cost functional to achieve a gate
that obeys experimental constraints and has a high simulated fidelity,
we also want to make the gate robust to noise that affects the experimental
fidelity of the gate. Decoherence of the quantum state due to external noise
is typically modeled by two phenomena: longitudinal relaxation and pure dephasing.
They are modeled using their $1/e$ decay times $T_{1}$ and $T_{\phi}$ respectively
(see Appendix).
The main contributions to longitudinal relaxation in our
device are dielectric loss in the capacitor, resistive loss in the inductor,
and Purcell loss. The main contributions to pure dephasing in our
device are $1/f$ flux noise and decay via charge and flux coupling
to the control lines.

Dissipation to the thermal bath via longitudinal
relaxation is an irreversible process
that results in information loss.
Converesely, pure dephasing is a reversible process.
There is a tradeoff between the two decoherence processes. In the case of white
noise we have that the sum of the noise weights $W_{1}$ and $W_{\phi}$
is constant \cite{huang2020engineering}.
Our device achieves its best pure dephasing
protection at the flux frustration point
$T_{2e}(\Phi_{\textrm{ext}} = 0.5 \Phi_{0}) \sim 300 \mu\textrm{s}$
where the qubit frequency is first-order insensitive to changes in flux.
It becomes more succeptable to pure dephasing as the flux is tuned away from the flux
frustration point. Conversely, $T_{1}$ is at a minimum
around the flux frustration point $T_{1}(0.5 \Phi_{0}) = 0.315$ms,
and increases away from the flux frustration point
$T_{1}(0.43 \Phi_{0}) = 4.3$ms. Given the nature
of the decay processes and the tradeoff, we choose
to maximize the longitudinal relaxation time ($T_{1}$)
over the gate duration
and employ robust control techniques to mitigate
pure dephasing.
%% TODO: See fig. 3 in Helin paper

%% LOH
%% SECTION 4
\section{Robustness to $T_{1}$-type Noise}
(Strategy) We seek to minimize the probability
that the qubit relaxes longitudinally.
To this end we augment the state vector $x_{k}$ with the
longitudinal relaxation probability
\begin{align}
  P_{1}(t_{k}) = \int_{0}^{t_{k}}
  \gamma_{1}(\Phi_{\textrm{ext}}(t)) dt
\end{align}
where $\gamma_{1} = T_{1}^{-1}$. Setting the target
longitudinal relaxation probability to 0 results in
a quadratic cost at each knot point
of the form ${\lvert P_{1}(t_{k}) \rvert}^{2}$.

@@

%% APPENDIX 2
\section{Experiment}
We measure $T_{1}$ using the standard experiment
and $T_{2}$ using te Ramsey experiment. We fit with splines
and the data looks like fig. 3 in Helin's paper \cite{zhang2020universal}.
We measure $\omega_{q}$ and $\sigma_{\omega_{q}}$ using X method.
